---
title: "Exploratory Data Analysis with dplyr"
output: pdf_document
---

# Functions
* `pull()`
* `select()`
* `unique()`
* `distinct()`
* `n_distinct()`
* `sort()`
* `filter()`
* `count()`
* `group_by()`
* `summarise()` (or `summarize()`)
* `arrange()`
* `disc()`
* `summary()`

```{r}
library("dplyr")
```

`dplyr` package has a sample dataframe "storms".

```{r}
storms
```

We can check a documentation about "storms".

```{r}
?storms # help(stomrs) works too
```

### Exploratory Data Analysis
To get to know the data. First step of analysis.  
1. Check data structure: first and last 5 rows, summarize statistics (mean, sd, etc), number of column and row...  
2. Clean the data: deleting some rows, filtering, sorting, deleting or replacing missing values  
3. Visualize the data: histogram, scatter plot  

`pull()` extracts column as a vector.

```{r}
pull(storms, year)
```

`select()` extracts column as a table.

```{r}
select(storms, year)
```

`unique()` returns only unique values.

```{r}
unique(pull(storms, year))
```

```{r}
unique(select(storms, year))
```

`storms` has records during a 40-year period from 1975 to 2015.

```{r}
max(pull(storms, year)) - min(pull(storms, year))
```

Inspect the column `month`.

```{r}
unique(pull(storms, month))
```

`sort()` sorts the order.

```{r}
sort(unique(pull(storms, month))) # increasing order (default)
```

```{r}
sort(unique(pull(storms, month)), decreasing = TRUE) # decreasing order
```

`sort()` takes only a vector, not a table.

```{r}
# This returns an error.
# sort(unique(select(storms, month)))
```


`filter()` returns rows that meet a logical condition.

```{r}
filter(storms, year == 1975) # extracts rows that year is 1975
```

Create an object that year is 1975.

```{r}
storms75 <- filter(storms, year == 1975)
```

How many kinds of storms in 1975?

```{r}
unique(pull(storms75, name)) # 3 storms recorded in 1975
```

Another way to find how many kinds of storms in 1975 is `distinct()`. It returns a tibble (table) instead of a vector.

```{r}
distinct(storms75, name)
```

`distinct()` takes only a table, not a vector.

```{r}
# This returns an error.
# distinct(pull(storms75, name))
```

`n_distinct()` returns a number of unique values.

```{r}
n_distinct(select(storms75, wind)) 

unique(select(storms75, wind))
```

`count()` returns frequencies of each value.

```{r}
count(storms75, name)
```

`group_by()` groups values. It is often used with `summarise()` to compute a summary on the specified column(s).

```{r}
summarise(group_by(storms75, name), avg_wind = mean(wind), avg_pressure = mean(pressure))
```

`summarize()` returns the same result , same usage ("summarise" and "summarize" are the synonyms).

```{r}
summarize(group_by(storms75, name), avg_wind = mean(wind), avg_pressure = mean(pressure))
```

Store the output.

```{r}
avg_wind_pressure <- summarise(group_by(storms75, name), avg_wind = mean(wind), avg_pressure = mean(pressure))
avg_wind_pressure # ordered by name alphabetically
```

`arrange()` orders rows and returns an ordered table.

```{r}
arrange(avg_wind_pressure, avg_wind) # order by average wind in increasing order
```

`desc()` orders decreasingly.

```{r}
arrange(avg_wind_pressure, desc(avg_wind))
```

The difference between `sort()` and `arrange()` is that `sort()` takes a vector and returns a vector, while `arrange()` takes a table and returns a table.

Focus on Amy in 1975.

```{r}
amy75 <- filter(storms75, name == "Amy")
amy75
```

Inspect `amy75`.

```{r}
distinct(amy75, status)
```

```{r}
distinct(amy75, month)
```

```{r}
count(distinct(amy75, day)) # How many days was Amy active?
```

`summary()` shows the statistical summary.

```{r}
summary(amy75) # Full stats
```

```{r}
summary(select(amy75, wind)) # stats only for wind
```




### Exercises
1) Use "dplyr" functions/commands to create a table (e.g. tibble) storm_names_1980s containing the name and year of storms recorded during the 1980s (i.e. from 1980 to 1989).

```{r}
storm_names_1980s <- distinct(filter(storms, year >= 1980 & year <= 1989), name, year)
storm_names_1980s
```


2) Use "dplyr" functions/commands to create a table (e.g. tibble) storms_per_year containing the number of storms recorded in each year (i.e. counts or frequencies of storms in each year). This table should contain two columns: year values in the first column, and number of storms in the second column.

```{r}
storms_per_year <- count(storms, year)
storms_per_year
```


3) Use "dplyr" functions/commands to create a table (e.g. tibble) storm_records_per_year containing three columns: 1) name of storm, 2) year of storm, and 3) count for number of records (of the corresponding storm).

```{r}
storm_records_per_year <- count(storms, name, year)
storm_records_per_year
```


4) Use "dplyr" functions/commands to display the different (unique) types of storm status.

```{r}
# they are all the same (but the last one returns a vector)
distinct(storms, status)

unique(select(storms, status))

unique(pull(storms, status))
```


5) Use "dplyr" functions/commands to display the different types of storm categories.

```{r}
distinct(storms, category)
```


6) Use "dplyr" functions/commands to create a table (e.g. tibble) storms_categ5 containing the name and year of those storms of category 5.

```{r}
storms_categ5 <- select(filter(storms, category == 5), name, year)
storms_categ5
```


7) Use "dplyr" functions/commands to display a table showing the status, avg_pressure (average pressure), and avg_wind (average wind speed), for each type of storm category. This table should contain four columns: 1) category, 2) status, 3) avg_pressure, and 4) avg_wind.

```{r}
summarise(group_by(storms, category, status), avg_pressure = mean(pressure), avg_wind = mean(wind))
```


8) Use "dplyr" functions/commands to create a table (e.g. tibble) max_wind_per_storm containing three columns: 1) year of storm, 2) name of storm, and 3) max_wind maximum wind speed record (for that storm).

```{r}
max_wind_per_storm <- summarise(group_by(storms, year, name), max_wind = max(wind))
max_wind_per_storm
```


9) Use "dplyr" functions/commands to create a table (e.g. tibble) max_wind_per_year containing three columns: 1) year of storm, 2) name of storm, and 3) wind maximum wind speed record (for that year). Arrange rows by wind speed in decreasing order.

```{r}
max_wind_per_year <- arrange(summarise(group_by(storms, year), 
                                       name = name[which.max(wind)], 
                                       max_wind = max(wind)), 
                             desc(max_wind))
max_wind_per_year
```

