---
title: "Factors"
output: pdf_document
---
**factor: **categorical variables

```{r}
position <- c('SG', 'PG', 'PF', 'SF', 'C')
```

Since factor $\neq$ character, the elements don't have "". `Levels` shows all categories.

```{r}
position_factor <- factor(position)
position_factor
```

`str()` checks a data structure. Notice that the categories are assigned integers.

```{r}
class(position)
class(position_factor)

str(position_factor)
```

So its type is `integer`.

```{r}
typeof(position_factor)
```

Factor can be treated like a vector in manipulation.

```{r}
position <- factor(position)

position[position == "SG"] # extract elemets whose value is "SG"
position[position == "spectator"] # a factor vector of length 0
position[1:3] # extract first 3 elements
position[rep(2, 4)] # extract the second element 4 times
```

```{r}
x <- c("c", "b", "a", "b", "c", "b")
y <- factor(x)
y
```

```{r}
y[3]
y[y == "sun"]
length(y)
sort(y)
unique(y)
```

But factor is not a vector. `attributes()` returns all levels (categories) and class.

```{r}
str(y)
is.vector(y) # factor is different from vector
attributes(y) # shows levels and class
```

# Qualitative Variables
- Nominal: no orders.
  - city has no orders.
  ```
  cities <- c("SF", "Berkeley", "SJ", "SJ", "SF", "SF")
  ```
- Ordinal: with orders.
  - rating has orders. It can be sorted like "bad" < "average" < "good".
  ```
  rates <- c("good", "bad", "bad", "good", "average", "good", "average", "good")
  ```

```{r}
rates <- c("good", "bad", "bad", "good", "average", "good", "average", "good")

# not ordered
factor(rates)
```

Notice that `Levels` is sorted alphabetically as default. Set `levels` to change the order.

```{r}
# ordered levels
rating <- factor(rates, 
                 levels = c("bad", "average", "good"))

rating
```

To give a quantative relationship, set `ordered`.

```{r}
# ordered levels with "<"
ordered_rating <- factor(rates, 
                  levels = c("bad", "average", "good"), 
                  ordered = TRUE)
ordered_rating
```

