---
title: "Geographical Maps with ggplot2"
output: pdf_document
---
```{r}
# load dplyr package
library(dplyr)
library(ggplot2)
```

```{r}
storms75 <- filter(storms, year == 1975)
head(storms75)
```

```{r}
ggplot(storms75, aes(x = long, y = lat)) + geom_point(aes(color = name))

# alternative ways to write equivalent commands
# ggplot(data = storms75) +  geom_point(aes(x = long, y = lat, color = name))
# ggplot() +  geom_point(data = storms75, aes(x = long, y = lat, color = name))
```

But where is it? We need an image of world map.

# Using `maps` package

```{r}
# install maps package
# install.packages("maps")

# load maps package
library(maps)
```

`map()` plots geographical maps.

```{r}
# world map
map("world")

# usa map
map("usa")

# usa states
map("state")

# usa counties
map("county")
```

To use `ggplot2` to map map objects, first we need to convert a map object to a dataframe object.

```{r}
world_map <- map_data("world")
head(world_map)

# check if world_map is really a dataframe object
class(world_map)
```

Use `geom_polygon()` to draw a map. `theme_bw()` changes the map to black-white.

```{r}
ggplot() + 
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group))

# use theme_bw()
# the background changes to white
ggplot() + 
  geom_polygon(data = world_map, aes(x = long, y = lat, group = group)) + 
  theme_bw()
```

```{r}
# store the ggplot to a variable
gg_world <- ggplot() + 
  geom_polygon(data = world_map, 
               aes(x = long, y = lat, group = group), 
               fill = "gray95", colour = "gray70", size = 0.2) +
  theme_bw()

gg_world
```

Let's add the points of the storm records. Use `xlim()` and `ylim()` to specify the range of longitude and latitude for zooming-in.

```{r}
gg_world + geom_point(data = storms75, aes(x = long, y = lat, color = name))

# zoom-in
gg_world + geom_point(data = storms75, aes(x = long, y = lat, color = name)) +
  xlim(c(-150, 0)) +
  ylim(c(0, 50))

# add wind speed feature by setting size argument
gg_world + 
  geom_point(data = storms75, 
             aes(x = long, y = lat, color = name, size = wind), alpha = 0.5) +
  xlim(c(-150, 0)) +
  ylim(c(0, 50))
```

`geom_path()` acts like `geom_point()`. It connects points but it's more like a "path" of the points.

```{r}
# similar result
gg_world + 
  geom_path(data = storms75, 
            aes(x = long, y = lat, color = name, size = wind), 
            alpha = 0.5, lineend = "round") +
  xlim(c(-150, 0)) +
  ylim(c(0, 50))
```

See the difference between `geom_line()` and `geom_path()` Both connects points but there's a big difference.

```{r}
# just an example
doris75 <- filter(storms75, name == "Doris")

# geom_path()
ggplot(data = doris75, aes(x = long, y = lat)) + geom_path()

# geom_line()
ggplot(data = doris75, aes(x = long, y = lat)) + geom_line()
```

# Using `rnaturalearth` package

```{r}
# install rnaturalearth package
# install.packages("rnaturalearth)
# may need additional packages
# install.packages(c("rnaturalearthdata", "rgeos"))

# load rnaturalearth package
library(rnaturalearth)

# library(rnaturalearthdata)
# library(rgeos)
```


`ne_countries()` gets world map polygons.

```{r}
world_df <- ne_countries(scale = "medium", returnclass = "sf")
class(world_df)
```

`geom_sf()` visualizes simple feature objects `sf`,

```{r}
ggplot(data = world_df) + geom_sf() + theme_bw()
```

We can set specify the region using `coord_sf()`, instead of using `xlim()` and `ylim()`.

```{r}
gg_world2 <- ggplot(data = world_df) + 
  geom_sf() + 
  coord_sf(xlim = c(-150, 0), y = c(0, 90), expand = TRUE) + 
  theme_bw()

gg_world2 + 
  geom_path(data = storms75, 
            aes(x = long, y = lat, color = name), 
            lineend = "round", size = 2, alpha = 0.5)
```

Facet by year from 1975 to 1980.

```{r}
storms_75_80 <- filter(storms, year %in% 1975:1980)

gg_world + geom_path(data = storms_75_80, 
                     aes(x = long, y = lat, group = name), 
                     lineend = "round") + 
  xlim(c(-150, 0)) + ylim(c(0, 90)) + 
  facet_wrap((~ year))
```

# Exercises
1) Filter storms in the 1980’s decade (1980 - 1989) and make a plot, with facets by month.year`

```{r}
storms80s <- filter(storms, year %in% 1980:1989)
head(storms80s)
```

Which year seems to have the largest number of storms?
Which year seems to have the smallest number of storms?
Does it seem to be a pattern (e.g. increasing number of storms over years)?

```{r}
num_of_storms80s <- distinct(group_by(storms80s, year, name), year)
num_of_storms80s

# by the bar plot, 1989 had the largest number of storms
# and 1987 had the smallest number of storms
ggplot(data = num_of_storms80s, aes(x = factor(year))) + # factor year to set label with only integers
  geom_bar()
```

The number of storms increased since 1980, but there seems no particular pattern over the period.


2) Take the previous data, storms in the 1980’s decade, and make a plot, with but this time with facets by month.

What is the most active month?
What is the least active month?
Are there months without any storms?

```{r}
num_storms_month <- distinct(group_by(storms80s, year, name, month), month)
num_storms_month
ggplot(data = num_storms_month, aes(x = month)) + geom_bar()

# facet by month but also year
num_storms_year_month <- distinct(group_by(storms80s, year, name, month), year, month)
num_storms_year_month
ggplot(data = num_storms_year_month, aes(x = factor(year))) + 
  geom_bar() + 
  facet_wrap(~ month) + 
  theme(axis.text.x = element_text(size=6)) # avoid overlapping labels
```

The most active month is September and least active month is December. No storms: January through May.