---
title: "Lists"
output: pdf_document
---

# Lists
- Atomic Structures
  - 1D: Vector: Only one type of data per vector
  - 2D: Matrix
  - nD: Array
- Non-Atomic Structures
  - 1D: List: Can contain any other type of data structure
  - 2D: Data Frame
  
# Create a list: `list()`

```{r}
lst <- list(1:3, matrix(1:9, 3, 3), list(1:2, c(TRUE, FALSE), c("a", "b")))
lst
```

To name each element of a list, 

```{r}
lst <- list("elem1" = 1:3, 
            "elem2" = matrix(1:9, 3, 3), 
            "elem3" = list(1:2, c(TRUE, FALSE), c("a", "b")))
lst
```

# Access to Elements of a List
- Single Brackets: Access to elements of a list, but not access to its object.

```{r}
lst[1]
```

```{r}
lst[2]
```

```{r}
lst[3]
```

- Double Brackets: Access to an object of an element of a list.

```{r}
lst[[1]]
```

```{r}
lst[[3]]
```

Combine single brackets and double brackets to access to a specific element of an object of an element of a list.

```{r}
lst[[1]][2]
```

```{r}
lst[[3]][2]
```

```{r}
lst[[3]][[2]][1]
```

- Dollar Signs: Only if a list has names for its elements.

```{r}
lst$elem2
```

```{r}
lst$elem2[2, 2]
```

# Adding Elements to a List
- `name_of_list[[new_index]] <- new_element`

```{r}
lst[[4]] <- c(T, F, T, F)
lst
```

- `name_of_list$new_name <- new_element`

```{r}
lst$elem5 <- 'nuevo'
lst
```

# Deleting Elements of a List
Use `NULL`.  
- `name_of_list[[index]] <- NULL`

```{r}
lst[[4]] <- NULL
lst
```

- `name_of_list$name_of_elem <- NULL`

```{r}
lst$elem5 <- NULL
lst
```

# Exercises
1) How would you create a list with your first name, middle name, and last name? For example, something like:

```
$first
[1] "Gaston"

$middle
NULL

$last
[1] "Sanchez"
```

```{r}
list("first" = "Gaston", "middle" = NULL, "last" = "Sanchez")
```

2) Consider an R list `student` containing the following elements:

```
$name
[1] "Luke Skywalker"

$major_minor
              major               minor 
     "jedi studies" "imperial policies" 

$gpa
[1] 4

$grades
           course score
1       force-101   9.3
2    light-sabers  10.0
3 jedi-literature   8.5
```

What is the output of the following commands? Try to guess the answer without running the code.

a. `student$grades$semester <- 4`

b. `sum(student[[2]] == "sith philosophy")`

c. `student["sid"] <- as.integer("123456")`

d. `mean(student[[4]][1:3,2], na.rm = TRUE)`

e. `student[[4]] <- student$grades[c(FALSE, TRUE, TRUE), ]`

```{r}
student <- list("name" = "Luke Skywalker", 
                "major_minor" = c("jedi studies", "imperial policies"), 
                "gpa" = 4, 
                "grades" = data.frame("course" = c("force-101", "light-sabers", "jedi-literature"), 
                                      "score" = c(9.3, 10.0, 8.5)))
student
```

```{r}
# a 
student$grades$semester <- 4
student
```

```{r}
# b
sum(student[[2]] == "sith philosophy")
```

```{r}
# c
student["sid"] <- as.integer("123456")
student
```

```{r}
# d
mean(student[[4]][1:3,2], na.rm = TRUE)
```

```{r}
# e
student[[4]] <- student$grades[c(FALSE, TRUE, TRUE), ]
student
```

