---
title: "Matrices and Arrays"
output: pdf_document
---
```{r}
library(dplyr)
```

Vector: 1-dimensional object  
Matrix: 2-dimensional object  
Array: n-dimensional object  

# Making Matrix and Array
- ` dim()`
- `matrix()`
- `array()`  
`matrix()` and `array()` are commonly used.

```{r}
x <- 1:8

# 2x4
dim(x) <- c(2, 4)
x
```

```{r}
x <- 1:8

# 2x2x2
dim(x) <- c(2, 2, 2)
x
```

```{r}
# 3x4
matrix(1:12, nrow = 3, ncol = 4)
```

```{r}
# 2x6
matrix(1:12, nrow = 2, ncol = 6, 
       byrow = TRUE, # fill elements from left to right
       dimnames = list(c("A", "B"), c("col1", "col2", "col3", "col4", "col5", "col6"))) # labels
```

```{r}
# 2x2x2
array(1:12, dim = c(2, 2, 3))
```

They will give an error if the number of elements in vector doesn't match the number of cells.

# Extract Elements of Matrices

- `name_of_matrix[row_index, col_index]`: specific cell.
- `name_of_matrix[row_index,]`: specific row.
- `name_of_matrix[, col_index]`: specific column.

```{r}
mat <- matrix(1:12, nrow = 3, ncol = 4)
mat
```

```{r}
# first row, second column
mat[1, 2]
```

```{r}
# second row
mat[2,]
```

```{r}
# third column
mat[, 3]
```

```{r}
# odd rows, even columns
mat[seq(1, 3, by = 2), seq(2, 4, by = 2)]
```

```{r}
# no second row, no first column
mat[-2, -1]
```

# Binding Rows and Columns

- `rbind()`: binds rows.
- `cbind()`: binds columns.

```{r}
mat2 <- matrix(1:8, nrow = 2, ncol = 4)
mat2
```

```{r}
# combine two matrices by row
rbind(mat, mat2)
```

```{r}
mat3 <- matrix(1:6, nrow = 3, ncol =2)
mat3
```


```{r}
# combine two matrices by column
cbind(mat, mat3)
```

# Mathematical Operation on Matrices

1. Transpose $X^T$: `t()`  
$X_{n \times m} \Rightarrow X_{m \times n}$

```{r}
# mat is 3x4 matrix
# its transpose is 4x3 matrix
t(mat)
```

2. Addition  
Matrices must be the same dimension.  
$\begin{bmatrix} 1 & 3 & 5 \\ 2 & 4 & 6 \end{bmatrix} + \begin{bmatrix} 7 & 9 & 8 \\ 8 & 7 & 9 \end{bmatrix} = \begin{bmatrix} 8 & 12 & 13\\ 10 & 11 & 15\end{bmatrix}$

```{r}
A <- matrix(1:6, 2, 3)
B <- matrix(7:9, 2, 3)

A + B
```

3. Scalar Multiplication  

```{r}
1/2 * A
```

4. Matrices Multiplication: `%*%`
$X_{a \times b} \times X_{b \times c} = X_{a \times c}$ 

```{r}
C <- matrix(7:9, 3, 2)
A %*% C
```

5. Multiplication by Elements: `*`
Matrices must be the same dimension.  

```{r}
A * A
```

6. Cross Product $X^TY$ or $XY^T$: `crossprod()`, `tcrossprod()`  

```{r}
# t(A) %*% A
crossprod(A, A)
```

```{r}
# A %*% t(A)
tcrossprod(A, A)
```

7. Linear Combination  
Multiplication by vector and matrix.  

```{r}
X <- matrix(1:12, 3, 4)
a <- seq(0.25, 1, by = 0.25)
b <- 1:3

X %*% a
```

```{r}
b %*% X
```

# Other Mathematical Functions of Matrices
1. `det()`: Find determinant.

```{r}
X2 <- rbind(c(-1,4,2), c(3,-2,1), c(1,1,1))
det(X2)
```

2. `diag()`: Extract or replace the diagonal elements.

```{r}
# 3x3 identity matrix
diag(3)
```

```{r}
# 3x3 diagonal elements 1, 2, 3
diag(1:3)
```

```{r}
# change diagonal elements to 5
I <- diag(3)
diag(I) <- 5
I
```

3. `solve()`: Solve the system of equations.  

```{r}
X3 <- rbind(c(-1,4,2), c(3,-2,1), c(1,1,1))
c <- c(1,2,-1)

solve(X3, c)
```

4. `eigen()`: Eigen decomposition.

```{r}
v <- matrix(c(8, 4, 1, 5), 2, 2)
v
eigen(v)
```

5. `svd()`: singular value decomposition. 

6. `qr()`: QR decomposition.

7. `chol()`: Choleski decomposition.

# Exercises
1) Given the following vector `h`p, how would you create the matrix displayed below?

```{r}
# vector of names
hp <- c("Harry", "Potter", "Ron", "Weasley", "Hermione", "Granger")
```

```
     [,1]     [,2]      
[1,] "Harry"  "Weasley"   
[2,] "Potter" "Hermione"  
[3,] "Ron"    "Granger"   
```

```{r}
matrix(hp, 3, 2)
```

2) Given the following vector `sw`, how would you create the matrix displayed below?

```{r}
# vector of names
sw <- c("Luke", "Skywalker", "Leia", "Organa", "Han", "Solo")
```

```
     [,1]   [,2]         
[1,] "Luke" "Skywalker"  
[2,] "Leia" "Organa"     
[3,] "Han"  "Solo"       
```

```{r}
matrix(sw, 3, 2, byrow = TRUE)
```

3) Consider the following vectors `a1`, `a2`, `a3`:

```{r}
a1 <- c(2, 3, 6, 7, 10)
a2 <- c(1.88, 2.05, 1.70, 1.60, 1.78)
a3 <- c(80, 90, 70, 50, 75)
```

Column-bind the vectors a1, a2, a3 to form the following matrix `M`:

```
#>   a1   a2 a3  
#> 1  2 1.88 80  
#> 2  3 2.05 90  
#> 3  6 1.70 70  
#> 4  7 1.60 50  
#> 5 10 1.78 75  
```
```{r}
M <- cbind(a1, a2, a3)
rownames(M) <- 1:5
M
```

4) Consider the following vectors `b1`, `b2`, `b3`:

```{r}
b1 <- c(1, 4, 5, 8, 9)
b2 <- c(1.22, 1.05, 3.60, 0.40, 2.54)
b3 <- c(20, 40, 30, 80, 100)
```

Row-bind the vectors b1, b2, b3 to form the following matrix `M2`:

```
#>        1     2    3    4      5  
#> b1  1.00  4.00  5.0  8.0   9.00  
#> b2  1.22  1.05  3.6  0.4   2.54  
#> b3 20.00 40.00 30.0 80.0 100.00  
```

```{r}
M2 <- rbind(b1, b2, b3)
colnames(M2) <- 1:5
M2
```

5) With matrices `M` and `M2` created above, use the matrix-multiplication operator `%*%` and the transpose function `t()` to compute the matrix products:

a. MM2

```{r}
M %*% M2
```

b. M2M

```{r}
M2 %*% M
```

c. $M^TM2^T$

```{r}
t(M) %*% t(M2)
```

d. $M2^TM^T$

```{r}
t(M2) %*% t(M)
```

