---
title: "Regular Expressions"
output: pdf_document
---

```{r}
library(stringi)
library(stringr)
library(dplyr)
```

**Regular expression** also known as **Regex** is a tool for describing patterns in strings. The package associated with regex in R are stringr and stringi. Regex is also commonly used in other languages with perhaps slight syntatic changes.

*stringr*: character manipulation.
*stringi*: character string processing facilities.

## 1) `stringi` package
`str_detect()` helps detect string patterns in a string. Ensure you have installed stringi and stringr packages. 

```{r}
city <- "Los Angeles"
str_detect(city,"os")
```

```{r}
cities <- c("Los Angeles", "New York", "Atlanta", "New Delhi")
str_detect(cities, "New")
```

```{r}
# case sensitive
str_detect(cities, "new")
```

`str_extract()` helps extract a substring from a string and `str_locate()` helps with the start and end points of a pattern.

```{r}
str_extract(cities, "New")
```

```{r}
str_locate(cities, "New")
```

`str_sub()` returns a substring of a string given start and end position.

```{r}
# Atlanta has only 7 characters
# thus it returns "nta"
str_sub(cities, start = 5, end = 8)
```

## 2) Regular Expressions
### 2.1) Anchors

Anchors help us assert the position, i.e., the beginning or end of the string.

|Anchor |Description| Example| 
|----------|---------|---------|
| `^` | Matches any string starting with a substring | `^New` |
| `$` | Matches any string ending with a substring | `y$` |
|Exact Match `^$` | Matches the string that starts and ends with substring | `^Hi There$` |

### 2.2) Character Classes

Character classes match any characters given a class

| Character Class | Match |
|-----------------|-------|
| `[aeiou]` | Matches vowels |
| `[0-9]` | Matches digits |
| `[a-z]` | Matches lower case letters |
| `[A-Z]` | Matches upper case letters |
| `[a-zA-Z]` | Matches both lower and upper case letters |
| `[a-g]` | Matches characters a through g | 

### 2.3) Quantifiers

Quantifiers quantify the number of instances of a character, group or character class. Your quantifier should be placed after the character/group/character class that is being quantified.

|Quantifier|Description|
|----------|----------|
| `P*` | 0 or more instances of P|
| `P+` | 1 or more instances of P|
| `P?` | 0 or 1 instance of P|
| `P{m}` | Exactly m instances of P|
| `P{m,}` | At least m instances of P|
| `P{m,n}` | Between m and n instances of P|

### 2.4) Logical Operators

| Operator | Usage | Example | Explanation |
|----------|-------|---------|-------------|
| `^`  | Not | `[^A-Za-z0-9]` | Identifies characters that are not alphanumeric |
|` \|` | Or  | `(Apple\|Orange)` | Identifies cases that have Apple or Oranges or both |

```{r}
str_extract(cities, regex("ew"))
```

```{r}
str_extract(cities, regex("^New [a-zA-Z]*"))
```

```{r}
banks = c("Bank of America", "Bank of the West", "Citibank", 
          "TD Bank", "Bank of England", "People\'s United Bank")

#banks starting with "Bank"
str_detect(banks, regex("^Bank"))

#banks ending with "Bank" (note: citibank is False as b is lowercase)
str_detect(banks, regex("Bank$"))

#bank called "TD Bank" (exact match)
str_detect(banks, regex("^TD Bank$"))

# same as above
str_detect(banks, "TD Bank")
```

```{r}
#Note the whitespace after Z in the regular expression
str_extract(banks, regex("^Bank[a-zA-Z ]*"))

#Without the whitespace after Z, it will not give us the desired output
str_extract(banks, regex("^Bank[a-zA-Z]*"))
```

```{r}
str_extract(banks,regex("^[a-zA-Z ]{1,9}$"))
```

```{r}
print(banks)

#Detect all bank names containing Bank or bank
str_detect(banks, regex("Bank|bank"))

#Detect all bank names containg special characters excluding whitespace
str_detect(banks, regex("[^A-z0-9 ]"))
```

```{r}
#List the names of first 10 nba players 
# with college name starting with "University"
nba = read.csv("nba2018-players.csv")

nba %>% 
  filter(str_detect(college, regex("^University"))) %>% head(10)

# List the names of colleges of nba players containing 
# special characters (don't consider whitespace and , 
# as special character here)
nba %>% 
  filter(str_detect(college, regex("[^A-Za-z, ]"))) %>% 
  select(college)
```

## 3) Using Apply 

Apply functions are useful when you want to apply a certain operation to all the rows of a list or dataframe. In the example below, we will only consider the case of a data frame. Note that this idea would be useful for your next lab. Here we apply `str_locate()` to locate area code from phone numbers. Using the output of `str_locate()`, we could then extract the area code.

```{r}
phone_num = c(
  "401-501-1111", 
  "(401)501-1111", 
  "401 501 1111", 
  "401-5011111", 
  "+408-501-1111")

code_pos = lapply(
  phone_num, 
  function(x) str_locate(x, pattern = regex('[0-9]{3}[- )]*')))

# Use substring() to extract the area code by sharing the start 
# and end position with the function as arguments.
code_pos = do.call(rbind, code_pos)
substring(phone_num, code_pos[,1], code_pos[,2]-1)
```

# Practice Problems

1. Print only those cities that start with `"New"` from `city`. Hint: Using `str_detect()`, we could create a mask (a vector of `TRUE` and `FALSE`) that could then be subsetted to get the city names starting with `"New"`.

```{r}
cities[str_detect(cities, "New")]
```

2. Using `str_match()` explained in cheatsheet, check if `city` has any matches for `Los Angeles`.

```{r}
str_match(cities, "Los Angeles")
```

3. Using `nba` dataframe, print names of Players containing `Marcus`

```{r}
# your code
nba %>% filter(str_detect(player, regex("Marcus"))) %>% select(player)
```

4. Using `nba` dataframe, print names of Universities that contain `California` or `Los Angeles` in them.

```{r}
# your code
nba %>% filter(str_detect(college, regex('California|Los Angeles'))) %>% select(college)
```

5. Consider the variable below `myVar`. Using str_detect and regular expression, detect those strings that contain at least one `z` and a maximum of three `z`. Your output should be `TRUE TRUE TRUE FALSE`

```{r}
myVar = c("bizarre", "bizzarre", "bizzzarrre", "bizzzzare")
```

```{r}
# your code
str_detect(myVar, regex("^[a-y]+z{1,3}[a-y]+"))
```

6. Consider the output of ls statement from command line. We have stored it in `file_names`, a vector containing file names. Detect file names that start with `STAT154` and have extension `.csv`
\
Expected Output: `TRUE FALSE FALSE TRUE FALSE TRUE`

```{r}
file_names = c(
  "STAT154nba.csv", 
  "test0102_STAT154.csv", 
  "STAT154myfile.csv.tmp", 
  "STAT154_lab.csv", 
  "STAT154_HW.pdf", 
  "STAT154_test0102.csv")
```

```{r}
# your code
str_detect(file_names, regex("^(STAT154)[a-zA-Z0-9_]+(.csv)$"))
```

7. For the variable `newVar`, identify strings that start with a number. Your output should be `TRUE FALSE TRUE TRUE FALSE`. 

```{r}
newVar = c(
  "1 Student(s)", 
  "None but 1 Student(s)", 
  "5 Students", 
  "120! Students", 
  "Two Students")
```

```{r}
# your code
str_detect(newVar, regex("^[0-9]"))
```

8. Detect phone numbers from a given list. Note that these won't have international codes. Your ouput for given test case should be `TRUE TRUE TRUE TRUE FALSE FALSE`
Hint: The last string in the vector has more than 10 digits.

```{r}
phone_num = c(
  "401-501-1111", 
  "(401)501-1111", 
  "401 501 1111", 
  "4015011111", 
  "+408-501-1111", 
  "40850211111")
```

```{r}
# your code
str_detect(phone_num, regex("([0-5()-[ ]]{6})([1]{4})"))
```

9. In the solution you gave above, can you think of cases where your expression would fail to detect an incorrect phone number? If yes, how could you improve it?

```{r}
# No code, but some text
#
# It would fail to detect an incorrect phone number 
# if the character length is longer or shorter than your average 10 character numbers.
```

10. In column of a dataframe, we have stored strings that tell us if someone likes orange juice, apple juice etc. Extract the names of students who like orange juice or oranges. Expected Output: Annie Harry
Hint: Use `str_detect` before using `str_extract()`

```{r}
myVar = c(
  "Annie likes Orange juice", "Sonny likes Apple juice", 
  "Katy dislikes Orange juice", "Harry likes Oranges", 
  "Charlie likes Apple juice", "Margo like Orange Pie")

df = data.frame(myVar)
```

```{r}
# your code
filtered <- df %>% filter(str_detect(myVar, regex(" likes Orange juice|( likes Oranges)$")))
str_extract(filtered$myVar, regex("^[A-Za-z]+"))
```

11. Using the idea above, from the `nba` dataframe, pull the last names of all 
NBA players. Find the frequencies of all the last names. _Hint:_ Use `str_locate()` and `substring()`. Eg: str_locate(nba$player, ...). Alternatively, you may choose to try this with lapply() but this is slightly tricky.

```{r}
# your code
head(str_sub(nba$player, str_locate(nba$player, pattern = regex('[^ ]+$'))), 10)
```

As the name suggests, it helps look around the string. `Look Arounds` indicate 
positions just like anchors, `$`, `^`, that we learnt in previous lab.

## 4) Look Aheads

The expression `A(?=B)` for look-ahead means "look for A, but match only if 
followed by B". There may be any pattern instead of A and B. 

### 4.1) Summary of Look-ahead

|Type|Syntax|Description|
|---|-----|-----------|
|Positive Look Ahead| `(?=pattern)`	| Lookahead	Asserts that what immediately follows the current position in the string is pattern|
|Negative Look Ahead| `(?!pattern)`	| Negative Lookahead	Asserts that what immediately follows the current position in the string is not pattern|

```{r}
myVar2 = "7 days of 10mm rainfall in Ohio and 2 nights of 15mm rainfall in NYC"

# extract all digits followed by "mm"
str_extract_all(myVar2, regex('[0-9]+(?=mm)'))

# extract all digits not followed by "mm"
str_extract_all(myVar2, regex('[0-9]+ (?!mm)'))
```

## 5) Look Behinds

Look Behind allows to match a pattern only if there’s something before it. This 
is contrary to lookahead which allows to assert for "what follows". The 
expression `(?<=B)A` matches A, but only if there’s B before it.

### 5.1) Summary of Look-behind

|Type|Syntax|Description|
|---|-----|-----------|
|Positive Look Behind| `(?<=pattern)` | Lookbehind	Asserts that what immediately precedes the current position in the string is pattern|
|Negative Look Behind| `(?<!pattern)` | Negative Lookbehind	Asserts that what immediately precedes the current position in the string is not pattern|

```{r}
myVar3 = '1 apple costs $1.50 in USA, 2 apples cost £3 elsewhere and 5€ in France'

# extract only prices
str_extract_all(myVar3, regex('((?<=[$£]{1})[0-9.]+)|([0-9.]+(?=[€]{1}))'))
```

# Prictice Problems
1. In these examples, lets try to extract the second university listed for every 
player in `sample_college` variable, a subset of the `nba` dataset we used earlier.

```{r}
nba = read.csv("nba2018-players.csv")

sample_college = nba[c(1,4,14,18,37,51,56,245,254:256,274:276,291),'college']
sample_college
```

In case our input string is `University of Memphis, University of Kansas`, we 
should get output as `University of Kansas`, i.e., print the second university 
name listed.

```{r}
str_extract(sample_college, regex("(?<=, ).*University.*"))
```

2. Using the expressions developed so far, write a function that returns a dataframe with two columns, last name of player and the name of second university that we extracted in last questions. The input to the function would be a subset of nba players. You only need to populate the function definition.

Hint: On running the above chunk, your output should be (for first five lines):

|  |Last_name  | Second_Univ|
|---|-----|-----|
|1 | McKinnie |      <NA>|
|2 |  DeRozan |      <NA>|
|3   |  Powell  |    <NA>|
|4   |   Nader   |   Iowa State University|
|5 |     Silas  |   Northern Illinois University|

```{r}
myfunc = function(df) {
  Last_name <- str_extract(df$player, regex("(?<= ).*"))
  Second_Univ <- str_extract(df$college, regex("(?<=, ).*University.*"))
  new_df <- data.frame(Last_name, Second_Univ)
  ### Write your code here and modify return statement
  return(head(new_df, 10))
}

myfunc(nba[c(1,4,14,18,37,51,56,245,254:256,274:276,291),])
```

3. In webscrapped content, one could retrieve desired content by looking for tags. In this case we would like to retrieve `firstHeading`. 

Example: `<h1 id="firstHeading" class="firstHeading" lang="en">University of California, Berkeley</h1>`

Here **University of Berkeley** is the content we need to retrieve, i.e. our `firstHeading`. First headings are always enclosed between 

(a) `<h1 id="firstHeading" class="firstHeading" lang="en">`
(b) `</h1>`

Write code to extract all occurances of First Headings in variable `text`. Use positive look ahead and positive look behind to capture the content between (a) and (b) given above.

Your output should be:

```
[[1]]
[1] "Yoshua Bengio"            "Turing Award"            
[3] "University of Manchester" "Chicken soup"
```

```{r}
text = '<h1 id="firstHeading" class="firstHeading" lang="en">University of California, Berkeley</h1><h1 id="firstHeading" class="firstHeading" lang="en">Chicken soup1<h1 id="firstHeading" class="firstHeading" lang="en">Yoshua Bengio</h1><h1 id="firstHeading" class="firstHeading" lang="en">Turing Award</h1><meta name="generator" content="MediaWiki 1.35.0-wmf.27"/><div id="siteSub" class="noprint">From Wikipedia, the free encyclopedia</div><div id="bodyContent" class="mw-body-content"><h1 id="firstHeading">Page as on Today</h1><h1 id="firstHeading" class="firstHeading" lang="en">University of Manchester</h1><h1 id="firstHeading" class="firstHeading" lang="en">Chicken soup</h1>'
```

```{r}
str_extract_all(text, regex('(?<="en">)[a-zA-Z ]*(?=</)'))
```

## 6) Some Metacharacters

|Expression|Description|
|----|------|
| `\\d` | match any digit (same as `[0-9]`)|
| `\\s` | match any whitespace (space, tab)|
| `\\t` | match only tab|
| `\\b` | match a word boundary |
| `\\A` | match the beginning of input|
| `\\Z` | match the end of input|

Metacharacters have a special meaning during pattern processing. Literals as we 
learnt in lectures are actual strings that we match. Lets look at some examples.

```{r}
myVar5 = "7 days of 10mm rainfall in Ohio and 2 nights of 15mm rainfall in NYC"

str_extract_all(myVar5, regex('\\d'))

str_extract_all(myVar5, regex('\\d\\d'))

str_extract_all(myVar5, regex('\\d+'))

str_extract_all(myVar5, regex('\\w'))

str_extract_all(myVar5, regex('\\w+'))
```

# Practice Problems

1. Write code to detect three digit area code The output for this case should be 
`TRUE TRUE FALSE FALSE`. Hint: Replace periods with solution `^[...]?\\d{...}[...]?$`

```{r}
test_case = c('401', '(401)', '4015', '+401')

str_detect(test_case, regex('^[(]?\\d{3}[)]?$'))
```

2. To the previous expression make changes to detect first 6 numbers. Your output here should be TRUE TRUE TRUE TRUE FALSE FALSE

```{r}
test_case2 = c('401 501', '401-501', '(401)501', '401501', '+401501', '4011501')

str_detect(test_case2, regex('^[(]?\\d{3}[) -]?\\d{3}$'))
```

3. Now consider the actual problem. Modify solution to previous question to find a pattern for detecting phone numbers. Your ouput for given test case should be TRUE TRUE TRUE TRUE FALSE FALSE

```{r}
phone_num = c(
  "401-501-1111", 
  "(401)501-1111", 
  "401 501 1111", 
  "4015011111", 
  "+408-501-1111", 
  "40850211111")

str_detect(phone_num, regex('^[(]?\\d{3}[) -]?\\d{3}[- ]?\\d{4}$'))
```

