---
title: "Vectors in R"
output: pdf_document
---

# What's Vector?
**Vector: ** a collection of same data type. It is called **Atomic structure**. Use `c()` to create a vector to combine multiple elements.     

Examples
```{r}
x <- 5 # a vector with length 1
x

1 + 2 # addition of vectors with length 1

c("a", "b", "c") # a vector of character
```

Most programming languages has a concept of scalar, but R doesn't.

# Name Elements

```{r}
x <- c("name" = "Vivi", "age" = 26, "sex" = "F")
x

# obtain just names
names(x)

# change names
names(x) <- c("NAaaaaME", "AGE", "SEX")
x

# name becomes NA if not defined
names(x) <- c("naaaaammmeeee")
x
```


# Accessing an Element of Vector
Use brackets `[]` after the variable.

```{r}
x <- 1:10 # 1 2 3 4 5 6 7 8 9 10
x[1]
x[3]
x[1:2]
x[-2]

x[x %% 2 == 0] # elements whose remainder is 0 when divided by 2

names(x) <- c("a", "b", "c", "d", "e", "f", "g", "h", "i", "j")
x["c"] # access by name

x[c(1, 3, 8)] # obtain elements of index 1, 3, and 8
```

# Adding Elements
1. `name of vector[new index] <- element`

```{r}
x <- 1:3
x
x[4] <- 100
x
x[6] <- -100
x # skip undefined index and the undefined element becomes NA
```

2. use `c()`

```{r}
x
x <- c(x, 22, Inf)
x

# with name
y <- c("a" = 11, "b" = 45, "c" = -1)
y[4] <- 30
y # 30 has no name
y<- c(y, 100, 3000)
y # 30, 100, and 3000 have no name
```

# Some Functions for Vector
- `length()`: returns a number of elements
- `sort()`: change the order of elements in increasing order. set `decreasing = TRUE` to order decreasingly
- `rev()`: reverse the order of elements
- `unique()`: returns only unique elements

```{r}
x <- c(1, 1, 2, 5, -9, 4, 2)
length(x)
sort(x)
sort(x, decreasing = TRUE)
rev(x)
unique(x)
x # they don't mutate the original
```

# Generate Sequences
1. `:`
2. `seq()`
3. `rep()`

```{r}
-2:5
10:1
-1:-5
-5:-1

seq(1, 10)
seq(from = 1, to = 10, by = 2)

rep(1, times = 3)
rep(c(1, 2), times = 3)
rep(1:2, each = 3)
rep(1:2, length.out = 5)
rep(c(3, 2, 1), times = 3, each = 2)

rep(seq(1, 5, 2), times = 2, each = 3)
```

# Vectorization
Applying a function on all elements of vector.

```{r}
x <- 1:3
log_x <- log(x)
log_x

x / 2
```

# Recycling
Recycle the elements of vector of shorter length when operating something with vector of different length.

```{r}
x <- 1:5
y <- 2:6 # same length as x
z <- 10 # vector of length 1

x + y # same length
x + z # different length
y - z # different length
x * y # same length
x * z # different length
```


# Data Types
There are 3 functions to check a data type:
* `typeof()` returns a basic type of the variable
  * vector (integer, double, character, etc), list, function, ...
* `mode()` returns a type of the element
  * numeric (integer, double), complex, logical, ...
* `class()` returns a type of class of the object
  * matrix, array, factor, ...
  * if class is not defined, it returns a basic type or mode.

- logical: `TRUE`, `FALSE`, `T`, `F`, `NA`

```{r}
typeof(TRUE)
typeof(NA)
```

- integer: to specify, add `L` next to a number; otherwise it is treated as "double" but it is rare to specify integer.

```{r}
typeof(1L)
typeof(1) # no "L" is treated as "double"

mode(1L)

class(1L)
```

- double: a number with decimal or real number.

```{r}
typeof(5)
typeof(-5)
typeof(-0.3)
typeof(1e-5) # e^-5
```

- complex: a complex number by adding "i".

```{r}
typeof(3 + 5i)
typeof(1i)
typeof(3 + 0i)
typeof(complex(re = 2, im = 3)) # 2 + 3i
```

- character: string(s). it is always a character if it is surrounded by `""`.

```{r}
typeof("a")
typeof("2020/09/10")
typeof("")
typeof(" ")
typeof("NA") # missing value
```

## Special Values
They usually remain the same after operation.  

- `NULL`: a vector of no elements of length 0.

```{r}
typeof(NULL)
```

- `NA`: a missing value.

```{r}
typeof(NA)
```

- `Inf`, `-Inf`: infinite number.

```{r}
typeof(Inf)
```

- `NaN`: not a number.

```{r}
typeof(NaN)
```

Example

```{r}
x <- c(1.0, NA, 3, 4.0)
x == NA # NA is applied to each element, and it returns just NA for each element
```

To check special values, use the following functions:  
- `is.null()`
- `is.na()`
- `is.finite()`
- `is.infinite()`
- `is.nan()`

# Coercion
1. Explicit Coercion  
change the data type of a vector, if possible.  

- `as.logical()`
- `as.integer()`
- `as.double()`
- `as.real()`
- `as.character()`

```{r}
x <- c("1", "2", "3.0")
typeof(x)
as.integer(x) # no "L" but integer
as.double(x)
as.complex(x)
x # they don't change the original vector

# the following gives an warning and returns NA if the element is not convertible.
y <- c("1", "abc", 3, "def", NA, Inf, "7.0")
as.integer(y)

# this won't work because of "implicit coercion"
as.logical(x) # just returns NA for each element
```

2. Implicit Coercion  
Hierarchy: **logical < numeric (integer < double) < complex < character**

```{r}
# different data types "double" and "complex" and "complex" dominates "double"
typeof(c(1, 1i))

z <- c(TRUE, FALSE, T, F, NA) # all elements are logical
z[2] <- -10
z # logical -> numeric
z[3] <- 0.5
z # integer -> double
z[4] <- 0.05
z # align decimal numbers
z[4] <- 5i
z # double -> complex
z[1] <- "-2"
z # complex -> character
z[5] <- 20
z # NA -> 20 (numeric) -> "20" (character)
```

# Exercises
1) Consider the following two vectors: x and y.
```
x <- c(2, 4, 6, 8, 10)
y <- c("a", "e", "i", "o", "u")
```
What is the output of the following R commands? (BTW: they are all valid commands). Try to answer these parts without running the code in R.

a)  `y[x/x]`

b)  `y[!(x > 5)]`

c)  `y[x < 10 & x != 2]`

d)  `y[x[-4][2]]`

e)  `y[as.logical(x)]`

f)  `y[6 - (x/2)]`

```{r}
x <- c(2, 4, 6, 7, 10)
y <- c("a", "e", "i", "o", "u")

# a
y[x/x]

# b
y[!(x > 5)]

# c
y[x < 10 & x != 2]

# d
y[x[-4][2]]

# e
y[as.logical(x)]

# f
y[6 - (x/2)]
```


2) Consider the following R code:
```
# peanut butter jelly sandwich
peanut <- TRUE
peanut[2] <- FALSE
yummy <- mean(peanut)
butter <- peanut + 1L
jelly <- tolower("JELLY")
sandwich <- c(peanut, butter, jelly)
```
What is the output of the following commands? Try to answer these parts without running the code in R.

a) `"jelly" != jelly`

b) `peanut & butter`

c) `typeof(yummy[peanut])`

d) `sandwich[2]`

e) `peanut[butter]`

f) `peanut %in% peanut`

g) `typeof(!yummy)`

h) `length(list(peanut, butter, as.factor(jelly)))`

```{r}
peanut <- TRUE
peanut[2] <- FALSE
yummy <- mean(peanut)
butter <- peanut + 1L
jelly <- tolower("JELLY")
sandwich <- c(peanut, butter, jelly)

# a
"jelly" != jelly

# b
peanut & butter

# c
typeof(yummy[peanut])

# d
sandwich[2]

# e
peanut[butter]

# f
peanut %in% peanut

# g
typeof(!yummy)

# h
length(list(peanut, butter, as.factor(jelly)))
```


3) Consider the following two vectors: x and y.
```
x <- c(1, 2, 3, 4, 5)
y <- c("a", "b", "c", "d", "e")
```
Match the following commands with their corresponding output. Try to answer these parts without running the code in R.
```
a)  y[x == 1]              ___  "a" "b" "c" "d" "e"

b)  y[x]                   ___  "e"

c)  y[x < 3]               ___  character(0)

d)  y[x/x]                 ___  "d"

e)  y[x[5]]                ___  "c" "d" "e"

f)  y['b']                 ___  NA

g)  y[0]                   ___  "a" "b"

h)  y[!(x < 3)]            ___  "c"

i)  y[x[-2][3]]            ___  "a"

j)  y[x[x[3]]]             ___  "a" "a" "a" "a" "a"
```

```{r}
x <- c(1, 2, 3, 4, 5)
y <- c("a", "b", "c", "d", "e")

# a
y[x == 1]

# b
y[x]

# c
y[x < 3]

# d
y[x/x]

# e
y[x[5]]

# f
y['b']

# g
y[0] # a character vector of length 0

# h
y[!(x < 3)]

# i
y[x[-2][3]]

# j
y[x[x[3]]]
```

4) Which command will fail to return the first five elements of a vector x? (assume x has more than 5 elements).

a) `x[1:5]`

b) `x[c(1,2,3,4,5)]`

c) `head(x, n = 5)`

d) `x[seq(1, 5)]`

e) `x(1:5)`

```{r}
# a
x[1:5]

# b
x[c(1, 2, 3, 4, 5)]

# c
head(x, n = 5)

# d
x[seq(1, 5)]

# e (answer)
# x(1:5) surrounding by () means applying a function
```

5) Explain the concept of atomic structures in R.

It is one type of the most fundamental data structure in R that can contain only one type of data. An atomic vector is either logical, integer, numeric, complex, character or raw and can have any attributes except a dimension attribute (like matrices). I.e., a `factor` is an atomic vector, but a `matrix` or `NULL` are not. In short, this is basically equivalent to `is.atomic(x) && !is.null(x) && is.null(dim(x))`. To check if it is an atomic vector, use `is.vector()`.

6) Explain the concept of vectorization a.k.a. vectorized operations.

Operations occur in parallel in certain R objects (i.e. They are computed element-by-element). This allows us to write code efficiently, concisely, and easily to read than in non-vectorized languages.